---
title: "Frangieh enrichment analysis"
output: pdf_document
date: "2023-04-21"
---

```{r, message = FALSE, echo = FALSE}
# load libraries and resolve conflicts
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(conflicted)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)

# analysis parameteres
N_NONZERO_CNTRL_THRESH <- 7
N_NONZERO_TRT_THRESH <- 7
q <- 0.1

# read the results
sceptre2_dir <- .get_config_path("LOCAL_SCEPTRE2_DATA_DIR")
method_results_fp <- paste0(
  sceptre2_dir, 
  "results/discovery_analyses/discovery_results_0423_processed.rds"
)
TF_targets_fp <- paste0(
  sceptre2_dir, 
  "results/discovery_analyses/TF_targets_frangieh_atacseq.rds"
)
method_results <- readRDS(method_results_fp) |> as_tibble()
TF_targets <- readRDS(TF_targets_fp) |> unique()

# join method results with TF targets, apply BH at level q
method_results <- method_results |> 
  filter(dataset_rename == "Frangieh Control Gene",
         n_control >= N_NONZERO_CNTRL_THRESH,
         n_treatment >= N_NONZERO_TRT_THRESH) |> 
  rename(gene = response_id, TF = grna_group) |> 
  inner_join(TF_targets, by = c("gene", "TF")) |>
  group_by(TF, Method) |>
  mutate(rejection = p.adjust(p_value, method = "fdr") <= q) |>
  ungroup()

# version of Fisher test that returns NA (rather than throwing an error) 
# in degenerate cases
my.fisher.test <- function(x, y){
  if(length(unique(x)) == 1 | length(unique(y)) == 1){
    list(p.value = NA, estimate = NA)
  } else{
    fisher.test(x,y)
  }
}

results <- method_results |> 
  group_by(TF, Method) |>
  summarise(num_rejections = sum(rejection),
            odds_ratio = my.fisher.test(target, rejection)$estimate,
            pvalue = my.fisher.test(target, rejection)$p.value, 
            .groups = "drop")
```

```{r, echo = FALSE}
my_cols <- c("Weissman Method" = "purple3",
             "Weissman Meth." = "purple3",
             "KS test" = "purple3",
             "Schraivogel Method" = "aquamarine4", 
             "Schraivogel Meth." = "aquamarine4",
             "MAST" = "aquamarine4",
             "Mimosca" = "palevioletred3",
             "MIMOSCA" = "palevioletred3",
             "Liscovitch Method" = "navy",
             "Liscovitch Meth." = "navy",
             "t-test" = "navy",
             "Seurat De" = "dodgerblue3",
             "Seurat De (w/ strict QC)" = "dodgerblue4",
             "Seurat De (w/o strict QC)" = "dodgerblue",
             "NB Reg (w/o covariates)" = "darkgoldenrod3",
             "NB Reg (w/ covariates)" = "darkgoldenrod4",
             "NB Regression" = "goldenrod3",
             "SCEPTRE" = "firebrick3",
             "Sceptre" = "firebrick3",
             "SCEPTRE (w/o covariates)" = "dodgerblue",
             "Truth" = "slategray4")

my_methods <- c("Liscovitch Method", "Schraivogel Method", "Mimosca", "Weissman Method", "Seurat De", "Sceptre", "Truth")
my_values <- my_cols[names(my_cols) %in% my_methods]
results <- results |>
  mutate(Method = factor(Method, levels = my_methods))
```

Below are the numbers of rejections, odds ratio, and p-values for each method and each TF. On the whole, the results are underwhelming. SCEPTRE makes a large number of discoveries only on MYC. For that TF, its p-value is not significant, but nearly so. On the other hand, the Liscovitch method does better on MYC. Note that the odds ratios and/or p-values for the some of the methods are infinite or undefined due to degeneracies. Often these have to do with having a very small number of discoveries, though for MIMOSCA, sometimes it finds all genes associated with the TF. 

```{r, fig.width = 9, fig.height = 8.5, echo = FALSE, warning = FALSE}
p_rejections <- results |>
  bind_rows(TF_targets |> 
    group_by(TF) |>
    summarise(num_rejections = sum(target)) |>
    mutate(Method = "Truth")
  ) |>
  mutate(Method = factor(Method, levels = my_methods)) |>
  ggplot(aes(x = num_rejections, y = TF, fill = Method)) +
  geom_col(position = "dodge", color = "black") + 
  scale_x_continuous(trans = "log1p", breaks = c(0, 1, 10, 100, 1000)) +
  scale_fill_manual(values = my_values) +
  theme(legend.position = "bottom",
        legend.title = element_blank()) + 
  labs(x = "Number of rejections", 
       y = "TF")

p_odds_ratios <- results |>
  ggplot(aes(x = odds_ratio, y = TF, fill = Method)) +
  geom_col(position = "dodge", color = "black") + 
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_continuous(trans = "log1p", breaks = c(0, 1, 2, 5, 10)) +
  scale_fill_manual(values = my_values) +
  theme(legend.position = "none") + 
  labs(x = "Odds ratio",
       y = NULL)

p_pvalues <- results |>
  ggplot(aes(x = pvalue, y = TF, fill = Method)) +
  geom_col(position = "dodge", color = "black") + 
  geom_vline(xintercept = 0.05, linetype = "dashed") +
  scale_x_continuous(trans = katlabutils::revlog_trans(),
                     breaks = c(1, 0.05, 0.01, 0.001)) +
  scale_fill_manual(values = my_values) +
  theme(legend.position = "") +
  labs(x = "p-value",
       y = NULL)

plot_grid(plot_grid(p_rejections + theme(legend.position = "none"), 
          p_odds_ratios, 
          p_pvalues, 
          nrow = 1),
          get_legend(p_rejections),
          ncol = 1,
          axis = "tlrb",
          rel_heights = c(1, 0.1))
```