---
title: 'Statistical debugging of `sceptre` part 2: calibration'
author: "Tim"
date: "2022-09-06"
output: html_document
---

```{r, message = FALSE, echo = FALSE}
# load packages
library(tidyverse)
library(katlabutils)
result_dir <- paste0(.get_config_path("LOCAL_SCEPTRE2_DATA_DIR"), "results/")

# load the auxilliary functions
funct_script <- paste0(.get_config_path("LOCAL_CODE_DIR"), "sceptre2-manuscript/writeups/undercover_results_eda/analyze_undercover_results_plot_functs.R")
source(funct_script)

# load results
# 1. Fisher exact test negative control results
# 2. Fisher exact test positive control results
# 3. effective sample size data frame (number of cells with nonzero expression per gRNA)
# 4. undercover result for group size = 1

# 1. 
fisher_exact_res <- readRDS(paste0(result_dir, "undercover_grna_analysis/fisher_exact_1.rds"))
# 2.
pc_res <- readRDS(paste0(result_dir, "positive_control_analysis/pc_results.rds"))
# 3.
sample_size_df <- readRDS(paste0(result_dir, "dataset_sample_sizes/n_nonzero_cells_per_grna.rds")) |>
  dplyr::mutate(dataset = dataset_concat,
                dataset_concat = NULL, paper = NULL, modality = NULL) |>
  dplyr::rename(grna_group = target, response_id = feature_id)
# 4.
undercover_res <- readRDS(paste0(result_dir, "undercover_grna_analysis/undercover_result_grp_size_1.rds")) |>
  dplyr::select(-clock_time, -max_ram) |>
  dplyr::filter(method %in% c("seurat_de", "liscovitch_method", "schraivogel_method", "weissman_method", "mimosca"))
# 5.
undercover_res_sceptre <- readRDS(paste0(result_dir, "undercover_grna_analysis/undercover_result_grp_size_1_sceptre.rds")) |>
  dplyr::select(-clock_time, -max_ram)
undercover_res <- rbind(undercover_res, undercover_res_sceptre)
```

```{r, message = FALSE, echo = FALSE}
# Combine the results of the Schraivogel datasets
replace_slash_w_underscore <- function(df) {
  df |> dplyr::mutate(dataset = gsub(pattern = "/",
                                replacement = "_",
                                fixed = TRUE,
                                x = dataset))
}

sample_size_df <-
  sample_size_df |>
  replace_slash_w_underscore() |>
  combine_schraivogel_enhancer_screens()

pc_res <- pc_res |>
  replace_slash_w_underscore() |>
  combine_schraivogel_enhancer_screens() |>
  update_dataset_names()

undercover_res <- undercover_res |>
  replace_slash_w_underscore() |>
  combine_schraivogel_enhancer_screens() |>
  update_dataset_names()
```

## Calibration of Fisher's exact test

I applied Fisher's exact test to analyze the negative control data within the context of the undercover pipeline. First, I plot the p-values on a transfored scale.

```{r, echo=FALSE, fig.height = 7, fig.width = 8, fig.align = "center"}
fisher_exact_res <- update_dataset_names(undercover_res = fisher_exact_res, TRUE)

p <- ggplot(data = fisher_exact_res,
             mapping = aes(y = p_value)) +
   stat_qq_points(ymin = 1e-10, size = 0.8) +
   facet_wrap(~dataset_rename_w_pairs, scales = "free", ncol = 3, labeller = label_wrap_gen(35)) +
   geom_abline(col = "darkred") +
   stat_qq_band() +
   theme_bw() +
   scale_x_continuous(trans = revlog_trans(10)) +
   scale_y_continuous(trans = revlog_trans(10)) +
   labs(x = "Expected quantile", y = "Observed quantile") +
   theme(legend.position = "bottom",
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.background = element_blank())
p
```

Overall, the results look quite good. Fisher's test exhibits good calibration far into the tail on all datasets except two: the Papalexi protein data and the Schraivogel perturb-seq data. Next, I plot the results on the untransformed scale.

```{r, echo=FALSE, fig.height = 7, fig.width = 8, fig.align = "center"}
p <- ggplot(data = fisher_exact_res,
             mapping = aes(y = p_value)) +
   stat_qq_points(ymin = 1e-10, size = 0.8) +
   facet_wrap(~dataset_rename_w_pairs, scales = "free", ncol = 3, labeller = label_wrap_gen(35)) +
   geom_abline(col = "darkred") +
   stat_qq_band() +
   theme_bw() +
   scale_x_reverse() +
   scale_y_reverse() +
   labs(x = "Expected quantile", y = "Observed quantile") +
   theme(legend.position = "bottom",
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.background = element_blank())
p
```

We see some conservative p-values in the bulk of the distribution.

# Positive control results

```{r, echo=FALSE,  fig.height = 7, fig.width = 8, warning=FALSE, message=FALSE}
n_bonf_reject <- compute_n_bonf_rejected(pc_res)
p <- make_n_rejected_pairs_plot(n_rejected_df = n_bonf_reject,
                                y_max = NULL,
                                scales = "free",
                                log_trans = FALSE)
p
```

The Fisher exact test lags somewhat behind the other methods on the metric of power; however, the Fisher exact test *does* seem capable of making discoveries on positive control data.

Now, we explore the connection between (effective) sample size and power (as quantified by the number of Bonferoni rejections on the positive control pairs).

```{r, echo = FALSE, message=FALSE}
# compute the real and effective sample size for each positive control pair
pc_res_pairs <- pc_res |>
  dplyr::select(grna_group, response_id, dataset) |>
  dplyr::distinct()

pc_res_pairs_plus_nt <- rbind(pc_res_pairs,
                              data.frame(grna_group = "non-targeting",
                                         response_id = pc_res_pairs$response_id,
                                         dataset = pc_res_pairs$dataset))
sample_size_df_sub <- dplyr::left_join(pc_res_pairs_plus_nt, sample_size_df,
                                       by = c("response_id", "grna_group", "dataset"))
samp_sizes <- apply(X = pc_res_pairs, MARGIN = 1, FUN = function(r) {
  grna_group <- as.character(r[["grna_group"]])
  response_id <- as.character(r[["response_id"]])
  dataset <- as.character(r[["dataset"]])
  # print(paste(grna_group, response_id, dataset))
  treatment_cell_sizes <- sample_size_df_sub |>
      dplyr::filter(response_id == !!response_id,
                    grna_group == !!grna_group,
                    dataset == !!dataset) |>
      summarize(s_n_nonzero = sum(n_nonzero_cells),
                s_n_cells = sum(n_cells))
  
  control_cell_sizes <- sample_size_df_sub |>
      dplyr::filter(response_id == !!response_id,
                    grna_group == "non-targeting",
                    dataset == !!dataset) |>
      summarize(s_n_nonzero = sum(n_nonzero_cells),
                s_n_cells = sum(n_cells))
  
    c(trt_n_nonzero = treatment_cell_sizes$s_n_nonzero,
               trt_n_cells = treatment_cell_sizes$s_n_cells,
               cntrl_n_nonzero = control_cell_sizes$s_n_nonzero,
               cntrl_n_cells = control_cell_sizes$s_n_cells)
}) |> t()
alpha <- 0.05
samp_sizes_2 <- cbind(pc_res_pairs, as.data.frame(samp_sizes)) |>
  dplyr::mutate(effective_sample_size = trt_n_nonzero + cntrl_n_nonzero)
pc_res_with_samp_sizes <- dplyr::left_join(pc_res, samp_sizes_2,
                                           by = c("grna_group", "response_id", "dataset")) |>
  dplyr::group_by(dataset, method) |>
  dplyr::mutate(reject = (p_value <= alpha/dplyr::n())) |>
  dplyr::ungroup() |>
  dplyr::mutate(lt = (p_value < 0.001))
```

Plot p-value vs. sample size.

```{r, echo = FALSE, message=FALSE, fig.height = 7, fig.width = 8}
# smallest effective sample size such that we reject
fisher_pc_res <- pc_res_with_samp_sizes |>
  dplyr::filter(method == "fisher_exact",
                dataset != "schraivogel_ground_truth_perturbseq_gene") |>
  dplyr::group_by(dataset) |>
  mutate(min_effective_sample_size = min(effective_sample_size[lt], na.rm = TRUE))

p <- ggplot(data = fisher_pc_res |> dplyr::filter(),
       mapping = aes(y = p_value, x = effective_sample_size, col = lt)) +
  geom_point(size = 0.8) +
  facet_wrap(dataset_rename_w_pairs ~ .,
             scales = "free_x",
             labeller = label_wrap_gen(35)) + theme_bw() +
  scale_color_manual(values = c("darkred", "darkblue")) +
  guides(color=guide_legend(title="p < 5e-4")) +
  geom_vline(mapping = aes(xintercept = min_effective_sample_size)) +
  xlab("Effective sample size (N cells with nonzero expression)")
p
```

```{r, echo = FALSE, message=FALSE}
fisher_pc_res |>
  select(dataset, min_effective_sample_size) |>
  distinct() |>
  rename("Min effective sample size to reject" = "min_effective_sample_size",
         "Dataset" = "dataset") |>
  knitr::kable()
```

What happens when we remove the \~30% of pairs with effective sample sizes that are too small? Does calibration become OK?

# Calibration after removal of pairs with small effective sample sizes

Effective sample size is simple to compute for the negative control analysis: for a given feature, simply count the number of negative control cells with nonzero expression.

```{r}
ntc_effective_samp_size <- sample_size_df |>
  filter(grna_group == "non-targeting") |>
  group_by(response_id, dataset) |>
  summarize(effective_samp_size = sum(n_nonzero_cells))
```

Many of the pairs in our dataset have an effective sample size of below 250.

```{r}
effective_samp_size_thresh <- 250
p <- ggplot(data = ntc_effective_samp_size,
       mapping = aes(x = effective_samp_size)) +
  geom_histogram() + facet_wrap(dataset~., scales ="free") +
  theme_bw() +
  xlab("Effective sample size") +
  ylab("Count") +
  scale_x_log10() + geom_vline(xintercept = effective_samp_size_thresh, col = "darkred")
```

First, here are the results on all pairs.

```{r, fig.height = 7, fig.width = 8}
undercover_res_w_samp_size <- dplyr::left_join(undercover_res,
                                               ntc_effective_samp_size, by = c("response_id", "dataset"))
make_trans_qq_plot(undercover_res_w_samp_size)
```

Next, here are the results on the pairs that pass the more stringent QC threshold.

```{r, fig.height = 7, fig.width = 8}
undercover_res_w_samp_size |>
  filter(effective_samp_size >= 250) |>
  make_trans_qq_plot()
```
