---
title: "Fast and accurate p-values for SCEPTRE: Preliminary Ideas"
author: "Gene"
date: "2023-01-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploration of resampling distributions

For convenience, all resampling distributions are stored in the folder `figures/`. There is one figure per dataset and effective sample sizes, each containing ten panels corresponding to ten pairs. Browsing these is illuminating, and I will summarize some of the main findings next. 

## Bumpy-ness of the resampling distributions

Many of the resampling distributions are bumpy, but interestingly, a given resampling distribution need not be bumpy throughout. Let $\mathcal B \subseteq [0,1]$ be be such that, informally, a given resampling distribution is bumpy at quantiles $q \in \mathcal B$. For example, Figure  shows the resampling distributions for the Schraivogel pairs with effective sample size 7. 

```{r schraivogel-7, fig.cap = "Schraivogel data (effective sample size = 7)", echo = FALSE}
knitr::include_graphics("figures/schraivogel_enhancer_screen_chr11_gene_eff_ss_7.pdf")
```

(and their tails in particular) have various levels of bumpy-ness. For each tail, we can categorize it as either bumpy or not bumpy. 

There are four distinct possibilities: 

Distributions can have 



